A comparison of design-based and model-based approaches for finite population spatial data.

### 2021-10-13 inst/manuscript/manuscript.Rmd
spell check and minor updates

While the Horvitz-Thompson estimator is unbiased for population means and totals, it is also important to quantify the uncertainty in these estimates. @horvitz1952generalization and @sen1953estimate provide variance estimators for $\hat{\tau}_{ht}$, but they have two drawbacks. First, these estimators rely on calculating $\pi_{ij}$, the probability that unit $i$ and unit $j$ are both in the sample -- this quantity can be challenging if not impossible to calculate analytically. Second, these estimators ignore the spatial locations of the units in the sampling frame. To address these two drawbacks simultaneously, @stevens2003variance proposed the local neighborhood variance estimator. The local neighborhood variance estimator does not rely on $\pi_{ij}$ and incorporates spatial locations -- for technical details see @stevens2003variance. @stevens2003variance show the local neigborhood variance estimator tends reduce $\text{Var}(\hat{\tau})$ compared to variance estimators ignoring spatial locations, yielding narrower confidence intervals for $\tau$.

@@

In addition to assuming the expectation of $\bm{\delta}$ equals $\mathbf{0}$, we also assume that there is spatial correlation in $\bm{\delta}$ that can be modeled using a covariance function. It is common to assume the covariance function is second-order stationary and isotropic [@cressie1993statistics], and that the spatial covariance decreases as the separation between population units increases. Many spatial covariance functions exist, but the primary function we use throughout the simulations and applications in this manuscript is the exponential covariance function: the $i,j$th entry for $\cov(\bm{\delta})$ is
\mbox{}
\begin{align}\label{equation:expcov}
\cov(\delta_i, \delta_j) = 
\begin{cases} 
\sigma^2_{ps}\exp(-h_{i,j}/\phi) & h_{i,j} > 0 \\
\sigma^2_{ps} + \sigma^2_n & h_{i,j} = 0
\end{cases}
,
\end{align}
where $\sigma^2_{ps}$ is the partial sill measuring coarse-scale (correlated) variability, $\sigma^2_{n}$ is the nugget measuring fine-scale (indepenent) variability, $\phi$ is the range parameter measuring the distance-decay rate of the covariance, and $h_{i,j}$ is the Euclidean distance between population units $i$ and $j$. Any spatial covariance function could be used in the place of the exponential, however, including functions that allow for non-stationarity or anisotropy [@chiles1999geostatistics, pp. 80-93].

@@

```{r, echo = FALSE, results = "asis"}
tab <- matrix(c("IRS-Design", "IRS-Model", "GRTS-Design", "GRTS-Model"),
       nrow = 2) %>%
  as.table()
row.names(tab) <- c("IRS", "GRTS")
xtab1 <- xtable(tab, col.names = c("Design", "Model"), caption = "\\label{tab:designanalysis} Types of Sampling Design and Analysis combinations considered in the simulation study. The rows give the two types of sampling designs while the columns give the two types of analyses.")
align(xtab1) <- "r|ll"
names(xtab1) <- c("IRS", "GRTS")
print(xtab1, comment = FALSE)
## I really want "Analysis" to appear as a row header and "Sampling Design" 
## to appear as a column header, but this isn't easy to figure out.
## I couldn't figure this out with xtable either so it might need to be
## made "by hand."
```

@@

The design-based and model-based approaches to inference are fundamentally different paradigms by which to select samples and analyze data. The design-based approach incorporates randomness through sampling to estimate a population parameter. The model-based approach incorporates randomness through distributional assumptions to predict the realized value of a random variable. Though these approaches have often been compared in the literature both from theoretical and analytical perspectives, our contribution lies in studying them in a spatial context while implementing spatially balanced sampling. Aside from the theoretical differences described, a few analytical findings were particularly notable: the design decision (GRTS vs IRS) seems much more important than the analysis decision (design-based vs model-based); Independent of the analysis approach, there is no reason to prefer IRS over GRTS for spatial data -- GRTS tends to perform at least as well as IRS when there is no spatial correlation increasingly than IRS as the strength of spatial correlation increases; the gap in relative efficiency between design-GRTS and model-GRTS widens as the strength of spatial correlation increases; and when the data are skewed, interval coverage for all approaches improves both as the sample size increases and as the strength of correlation increases.

There are several benefits and drawbacks of the design-based and model-based approaches for spatial data, some of which we have not yet discussed but are worthy of consideration in future research. The design-based approach relies on few assumptions, while the model-based approach relies on rigid distributional ones. The Horvitz-Thompson estimator of means and totals is unbiased and computationally efficient, while model-based estimation of covariance parameters can be computationally burdensome, especially for likelihood-based methods such as REML that rely on inverting a covariance matrix. The design-based approach also more naturally handles binary data, free from the more complicated logistic regression formulation commonly used to handle binary data in a model-based approach. The model-based approach, however, can quantify the relationship between covariates (predictor variables) and the response variable, something the design-based approach cannot do naturally. The model-based approach also yields estimated spatial covariance parameters, which help better understand the process of study. Model selection is also possible using model-based approaches and criteria such as likelihood ratio tests or AIC [@akaike1974new]. Model-based approaches are capable of more efficient small-area estimation than design-based approaches by leveraging distributional assumptions in areas with few observed sites. Mdoel-based approaches can also compute site-by-site predictions at unobserved locations and use them to construct informative visualizations. The benefits and drawbacks of both approaches, alongside our theoretical and analytical comparisons, should be heavily considered when choosing among them. This is especially true from an analysis perspective, as we found that using a spatially balanced sampling algorithm benefits both design-based and model-based analyses.

>>>>

While the Horvitz-Thompson estimator is unbiased for population means and totals, it is also important to quantify the uncertainty in these estimates. @horvitz1952generalization and @sen1953estimate provide variance estimators for $\hat{\tau}_{ht}$, but they have two drawbacks. First, these estimators rely on calculating $\pi_{ij}$, the probability that unit $i$ and unit $j$ are both in the sample -- this quantity can be challenging if not impossible to calculate analytically. Second, these estimators ignore the spatial locations of the units in the sampling frame. To address these two drawbacks simultaneously, @stevens2003variance proposed the local neighborhood variance estimator. The local neighborhood variance estimator does not rely on $\pi_{ij}$ and incorporates spatial locations -- for technical details see @stevens2003variance. @stevens2003variance show the local neighborhood variance estimator tends reduce $\text{Var}(\hat{\tau})$ compared to variance estimators ignoring spatial locations, yielding narrower confidence intervals for $\tau$.

@@

In addition to assuming the expectation of $\bm{\delta}$ equals $\mathbf{0}$, we also assume that there is spatial correlation in $\bm{\delta}$ that can be modeled using a covariance function. It is common to assume the covariance function is second-order stationary and isotropic [@cressie1993statistics], and that the spatial covariance decreases as the separation between population units increases. Many spatial covariance functions exist, but the primary function we use throughout the simulations and applications in this manuscript is the exponential covariance function: the $i,j$th entry for $\cov(\bm{\delta})$ is
\mbox{}
\begin{align}\label{equation:expcov}
\cov(\delta_i, \delta_j) = 
\begin{cases} 
\sigma^2_{ps}\exp(-h_{i,j}/\phi) & h_{i,j} > 0 \\
\sigma^2_{ps} + \sigma^2_n & h_{i,j} = 0
\end{cases}
,
\end{align}
where $\sigma^2_{ps}$ is the partial sill measuring coarse-scale (correlated) variability, $\sigma^2_{n}$ is the nugget measuring fine-scale (independent) variability, $\phi$ is the range parameter measuring the distance-decay rate of the covariance, and $h_{i,j}$ is the Euclidean distance between population units $i$ and $j$. Any spatial covariance function could be used in the place of the exponential, however, including functions that allow for non-stationarity or anisotropy [@chiles1999geostatistics, pp. 80-93].

@@

```{r, echo = FALSE, results = "asis"}
tab <- matrix(c("IRS-Design", "IRS-Model", "GRTS-Design", "GRTS-Model"),
       nrow = 2, byrow = TRUE) %>%
  as.table()
row.names(tab) <- c("IRS", "GRTS")
xtab1 <- xtable(tab, col.names = c("Design", "Model"), caption = "\\label{tab:designanalysis} Types of Sampling Design and Analysis combinations considered in the simulation study. The rows give the two types of sampling designs while the columns give the two types of analyses.")
align(xtab1) <- "r|ll"
names(xtab1) <- c("Design", "Model")
print(xtab1, comment = FALSE)
## I really want "Analysis" to appear as a row header and "Sampling Design" 
## to appear as a column header, but this isn't easy to figure out.
## I couldn't figure this out with xtable either so it might need to be
## made "by hand."
```

@@

The design-based and model-based approaches to inference are fundamentally different paradigms by which to select samples and analyze data. The design-based approach incorporates randomness through sampling to estimate a population parameter. The model-based approach incorporates randomness through distributional assumptions to predict the realized value of a random variable. Though these approaches have often been compared in the literature both from theoretical and analytical perspectives, our contribution lies in studying them in a spatial context while implementing spatially balanced sampling. Aside from the theoretical differences described, a few analytical findings were particularly notable: the design decision (GRTS vs IRS) seems much more important than the analysis decision (design-based vs model-based); Independent of the analysis approach, there is no reason to prefer IRS over GRTS for spatial data -- GRTS tends to perform at least as well as IRS when there is no spatial correlation increasingly than IRS as the strength of spatial correlation increases; the gap in relative efficiency between GRTS-design and GRTS-model widens as the strength of spatial correlation increases; and when the data are skewed, interval coverage for all approaches improves both as the sample size increases and as the strength of correlation increases.

There are several benefits and drawbacks of the design-based and model-based approaches for spatial data, some of which we have not yet discussed but are worthy of consideration in future research. The design-based approach relies on few assumptions, while the model-based approach relies on rigid distributional ones. The Horvitz-Thompson estimator of means and totals is unbiased and computationally efficient, while model-based estimation of covariance parameters can be computationally burdensome, especially for likelihood-based methods such as REML that rely on inverting a covariance matrix. The design-based approach also more naturally handles binary data, free from the more complicated logistic regression formulation commonly used to handle binary data in a model-based approach. The model-based approach, however, can quantify the relationship between covariates (predictor variables) and the response variable, something the design-based approach cannot do naturally. The model-based approach also yields estimated spatial covariance parameters, which help better understand the process of study. Model selection is also possible using model-based approaches and criteria such as likelihood ratio tests or AIC [@akaike1974new]. Model-based approaches are capable of more efficient small-area estimation than design-based approaches by leveraging distributional assumptions in areas with few observed sites. Model-based approaches can also compute site-by-site predictions at unobserved locations and use them to construct informative visualizations. The benefits and drawbacks of both approaches, alongside our theoretical and analytical comparisons, should be heavily considered when choosing among them. This is especially true from an analysis perspective, as we found that using a spatially balanced sampling algorithm benefits both design-based and model-based analyses.

### 2021-10-13 inst/manuscript/manuscript.tex
spell check and minor updates

While the Horvitz-Thompson estimator is unbiased for population means
and totals, it is also important to quantify the uncertainty in these
estimates. Horvitz and Thompson (1952) and Sen (1953) provide variance
estimators for \(\hat{\tau}_{ht}\), but they have two drawbacks. First,
these estimators rely on calculating \(\pi_{ij}\), the probability that
unit \(i\) and unit \(j\) are both in the sample -- this quantity can be
challenging if not impossible to calculate analytically. Second, these
estimators ignore the spatial locations of the units in the sampling
frame. To address these two drawbacks simultaneously, Stevens and Olsen
(2003) proposed the local neighborhood variance estimator. The local
neighborhood variance estimator does not rely on \(\pi_{ij}\) and
incorporates spatial locations -- for technical details see Stevens and
Olsen (2003). Stevens and Olsen (2003) show the local neigborhood
variance estimator tends reduce \(\text{Var}(\hat{\tau})\) compared to
variance estimators ignoring spatial locations, yielding narrower
confidence intervals for \(\tau\).

@@

In addition to assuming the expectation of \(\bm{\delta}\) equals
\(\mathbf{0}\), we also assume that there is spatial correlation in
\(\bm{\delta}\) that can be modeled using a covariance function. It is
common to assume the covariance function is second-order stationary and
isotropic (Cressie, 1993), and that the spatial covariance decreases as
the separation between population units increases. Many spatial
covariance functions exist, but the primary function we use throughout
the simulations and applications in this manuscript is the exponential
covariance function: the \(i,j\)th entry for
\(\mathop{\mathrm{{cov}}}(\bm{\delta})\) is \mbox{}
\begin{align}\label{equation:expcov}
\mathop{\mathrm{{cov}}}(\delta_i, \delta_j) = 
\begin{cases} 
\sigma^2_{ps}\exp(-h_{i,j}/\phi) & h_{i,j} > 0 \\
\sigma^2_{ps} + \sigma^2_n & h_{i,j} = 0
\end{cases}
,
\end{align} where \(\sigma^2_{ps}\) is the partial sill measuring
coarse-scale (correlated) variability, \(\sigma^2_{n}\) is the nugget
measuring fine-scale (indepenent) variability, \(\phi\) is the range
parameter measuring the distance-decay rate of the covariance, and
\(h_{i,j}\) is the Euclidean distance between population units \(i\) and
\(j\). Any spatial covariance function could be used in the place of the
exponential, however, including functions that allow for
non-stationarity or anisotropy (Chiles and Delfiner, 1999, pp. 80--93).

@@

\begin{table}[ht]
\centering
\begin{tabular}{r|ll}
  \hline
 & IRS & GRTS \\ 
  \hline
IRS & IRS-Design & GRTS-Design \\ 
  GRTS & IRS-Model & GRTS-Model \\ 
   \hline
\end{tabular}
\caption{\label{tab:designanalysis} Types of Sampling Design and Analysis combinations considered in the simulation study. The rows give the two types of sampling designs while the columns give the two types of analyses.} 
\end{table}

@@

The design-based and model-based approaches to inference are
fundamentally different paradigms by which to select samples and analyze
data. The design-based approach incorporates randomness through sampling
to estimate a population parameter. The model-based approach
incorporates randomness through distributional assumptions to predict
the realized value of a random variable. Though these approaches have
often been compared in the literature both from theoretical and
analytical perspectives, our contribution lies in studying them in a
spatial context while implementing spatially balanced sampling. Aside
from the theoretical differences described, a few analytical findings
were particularly notable: the design decision (GRTS vs IRS) seems much
more important than the analysis decision (design-based vs model-based);
Independent of the analysis approach, there is no reason to prefer IRS
over GRTS for spatial data -- GRTS tends to perform at least as well as
IRS when there is no spatial correlation increasingly than IRS as the
strength of spatial correlation increases; the gap in relative
efficiency between design-GRTS and model-GRTS widens as the strength of
spatial correlation increases; and when the data are skewed, interval
coverage for all approaches improves both as the sample size increases
and as the strength of correlation increases.

There are several benefits and drawbacks of the design-based and
model-based approaches for spatial data, some of which we have not yet
discussed but are worthy of consideration in future research. The
design-based approach relies on few assumptions, while the model-based
approach relies on rigid distributional ones. The Horvitz-Thompson
estimator of means and totals is unbiased and computationally efficient,
while model-based estimation of covariance parameters can be
computationally burdensome, especially for likelihood-based methods such
as REML that rely on inverting a covariance matrix. The design-based
approach also more naturally handles binary data, free from the more
complicated logistic regression formulation commonly used to handle
binary data in a model-based approach. The model-based approach,
however, can quantify the relationship between covariates (predictor
variables) and the response variable, something the design-based
approach cannot do naturally. The model-based approach also yields
estimated spatial covariance parameters, which help better understand
the process of study. Model selection is also possible using model-based
approaches and criteria such as likelihood ratio tests or AIC (Akaike,
1974). Model-based approaches are capable of more efficient small-area
estimation than design-based approaches by leveraging distributional
assumptions in areas with few observed sites. Mdoel-based approaches can
also compute site-by-site predictions at unobserved locations and use
them to construct informative visualizations. The benefits and drawbacks
of both approaches, alongside our theoretical and analytical
comparisons, should be heavily considered when choosing among them. This
is especially true from an analysis perspective, as we found that using
a spatially balanced sampling algorithm benefits both design-based and
model-based analyses.

>>>>

While the Horvitz-Thompson estimator is unbiased for population means
and totals, it is also important to quantify the uncertainty in these
estimates. Horvitz and Thompson (1952) and Sen (1953) provide variance
estimators for \(\hat{\tau}_{ht}\), but they have two drawbacks. First,
these estimators rely on calculating \(\pi_{ij}\), the probability that
unit \(i\) and unit \(j\) are both in the sample -- this quantity can be
challenging if not impossible to calculate analytically. Second, these
estimators ignore the spatial locations of the units in the sampling
frame. To address these two drawbacks simultaneously, Stevens and Olsen
(2003) proposed the local neighborhood variance estimator. The local
neighborhood variance estimator does not rely on \(\pi_{ij}\) and
incorporates spatial locations -- for technical details see Stevens and
Olsen (2003). Stevens and Olsen (2003) show the local neighborhood
variance estimator tends reduce \(\text{Var}(\hat{\tau})\) compared to
variance estimators ignoring spatial locations, yielding narrower
confidence intervals for \(\tau\).

@@

In addition to assuming the expectation of \(\bm{\delta}\) equals
\(\mathbf{0}\), we also assume that there is spatial correlation in
\(\bm{\delta}\) that can be modeled using a covariance function. It is
common to assume the covariance function is second-order stationary and
isotropic (Cressie, 1993), and that the spatial covariance decreases as
the separation between population units increases. Many spatial
covariance functions exist, but the primary function we use throughout
the simulations and applications in this manuscript is the exponential
covariance function: the \(i,j\)th entry for
\(\mathop{\mathrm{{cov}}}(\bm{\delta})\) is \mbox{}
\begin{align}\label{equation:expcov}
\mathop{\mathrm{{cov}}}(\delta_i, \delta_j) = 
\begin{cases} 
\sigma^2_{ps}\exp(-h_{i,j}/\phi) & h_{i,j} > 0 \\
\sigma^2_{ps} + \sigma^2_n & h_{i,j} = 0
\end{cases}
,
\end{align} where \(\sigma^2_{ps}\) is the partial sill measuring
coarse-scale (correlated) variability, \(\sigma^2_{n}\) is the nugget
measuring fine-scale (independent) variability, \(\phi\) is the range
parameter measuring the distance-decay rate of the covariance, and
\(h_{i,j}\) is the Euclidean distance between population units \(i\) and
\(j\). Any spatial covariance function could be used in the place of the
exponential, however, including functions that allow for
non-stationarity or anisotropy (Chiles and Delfiner, 1999, pp. 80--93).

@@

\begin{table}[ht]
\centering
\begin{tabular}{r|ll}
  \hline
 & Design & Model \\ 
  \hline
IRS & IRS-Design & IRS-Model \\ 
  GRTS & GRTS-Design & GRTS-Model \\ 
   \hline
\end{tabular}
\caption{\label{tab:designanalysis} Types of Sampling Design and Analysis combinations considered in the simulation study. The rows give the two types of sampling designs while the columns give the two types of analyses.} 
\end{table}

@@

The design-based and model-based approaches to inference are
fundamentally different paradigms by which to select samples and analyze
data. The design-based approach incorporates randomness through sampling
to estimate a population parameter. The model-based approach
incorporates randomness through distributional assumptions to predict
the realized value of a random variable. Though these approaches have
often been compared in the literature both from theoretical and
analytical perspectives, our contribution lies in studying them in a
spatial context while implementing spatially balanced sampling. Aside
from the theoretical differences described, a few analytical findings
were particularly notable: the design decision (GRTS vs IRS) seems much
more important than the analysis decision (design-based vs model-based);
Independent of the analysis approach, there is no reason to prefer IRS
over GRTS for spatial data -- GRTS tends to perform at least as well as
IRS when there is no spatial correlation increasingly than IRS as the
strength of spatial correlation increases; the gap in relative
efficiency between GRTS-design and GRTS-model widens as the strength of
spatial correlation increases; and when the data are skewed, interval
coverage for all approaches improves both as the sample size increases
and as the strength of correlation increases.

There are several benefits and drawbacks of the design-based and
model-based approaches for spatial data, some of which we have not yet
discussed but are worthy of consideration in future research. The
design-based approach relies on few assumptions, while the model-based
approach relies on rigid distributional ones. The Horvitz-Thompson
estimator of means and totals is unbiased and computationally efficient,
while model-based estimation of covariance parameters can be
computationally burdensome, especially for likelihood-based methods such
as REML that rely on inverting a covariance matrix. The design-based
approach also more naturally handles binary data, free from the more
complicated logistic regression formulation commonly used to handle
binary data in a model-based approach. The model-based approach,
however, can quantify the relationship between covariates (predictor
variables) and the response variable, something the design-based
approach cannot do naturally. The model-based approach also yields
estimated spatial covariance parameters, which help better understand
the process of study. Model selection is also possible using model-based
approaches and criteria such as likelihood ratio tests or AIC (Akaike,
1974). Model-based approaches are capable of more efficient small-area
estimation than design-based approaches by leveraging distributional
assumptions in areas with few observed sites. Model-based approaches can
also compute site-by-site predictions at unobserved locations and use
them to construct informative visualizations. The benefits and drawbacks
of both approaches, alongside our theoretical and analytical
comparisons, should be heavily considered when choosing among them. This
is especially true from an analysis perspective, as we found that using
a spatially balanced sampling algorithm benefits both design-based and
model-based analyses.
